The variability of Betelgeuse explained by surface convection

### 2024-04-17 main.tex
Add files via upload

Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days, and other shorter periods of 400 and 200 days. The origin of the LSP remains poorly known, some tried to explain the LSP by the lifetime of the large convective cells present at the surface of RSG (e.g \cite{stothers_giant_2010}) or by the magnetic field \citep{wood_long_2004}. The shorter periods of 400 and 200 days seem to have a different origin than the LSP. By studying a large sample of RSG, \cite{kiss_variability_2006} attributed the 400 days period to radial pulsation modes, the 200 days period being attributed to the first overtone \citep{joyce_standing_2020}, which was in agreement with the historical work of \cite{stothers_pulsation_1969}.\\

@@

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, it is speculated that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, that might be caused by a mass loss event and leading to this dimming \citep{dupree_great_2022}. Other theories such as a drop in temperature was also mentioned by \cite{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. In this event it was clear that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk.
%due to the presence of large quantities of dust.
Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period, while after the great dimming the main periods of Betelgeuse have shortened, oscillating between 97 days and 230 days \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon. 
Our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 400 and 200 days in  
observational proxies related to the convective activity.\\
%but unrelated to pulsations and which cannot be directly linked to brightness.
%Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
%require,

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes Q.pdf}
    \caption{Same as figure \ref{LS intensity} for stokes Q. }
    \label{LS Q}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes U.pdf}
    \caption{Same as figure \ref{LS intensity} for stokes U.}
    \label{LS U}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle linear polarization.pdf}
    \caption{Same as figure \ref{LS intensity} for linear polarization: $\sqrt{\mathrm{Q^2}+\mathrm{U^2}}$.}
    \label{LS linear polarization}
\end{figure}

@@

Figure \ref{LS Q} and \ref{LS U} show the Lomb-Scargle periodogram of both stokes $Q$ and $U$. It is clear that the periodogram is completely different 
from the intensity one. A strong signal is present in fig. \ref{LS Q} around 200 days. Concerning stokes U in fig. \ref{LS U}, it appears that the main period is around $0.003 \ \mathrm{days^{-1}}$ (330 days). Other period such as the one at 200 days or 250 days ($0.004 \ \mathrm{days^{-1}}$) are present but difficult to trust. From stokes $Q$ and $U$, we recover the 200 days period and the 330 days period is more or less close to the 400 days period found in the literature. 
Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of the total linear polarization of Betelgeuse: $\sqrt{\mathrm{Q^2+U^2}}$. 
Interestingly, the periodogram shows a strong power at both 330 days and 200 days, which are the period that we have found in both stokes $Q$ and $U$.

@@

Betelgeuse is observed in average every two or three weeks by the TBL, and hence
we are able to follow its surface activity through polarimetry imaging. This has  allowed in the past the estimation of  the size and the
 lifetime of the convective cells
at the surface. From these images, we computed a photo-center, which is sensitive to the size and the number of convective cells. 
An homogeneous star will have a photo-center displacement coinciding with barycenter of the star, whereas a star with one or two huge convective cells will have a
photo-center displacement more important, up to a few percent of the stellar radius in the case of red supergiants \citep{chiavassa_probing_2022}.

Since the photo-center is linked to surface convection, it is worth checking for periods in its 
dynamics over the 13 years of observations of Betelgeuse at the TBL. Such periods will not 
be completely unrelated to the ones presented in the previous section, but will capture 
other aspects of the phenomena at work.

@@

RSGs encounter mass loss event, where due to stellar winds, plasma is ejected in the interstellar medium. Those events have already been seen in RSG like $\mu$Cep, where \cite{lopez_ariste_height_2023} found an excess of linear polarization beyond the limb of the star and associated it to rising plumes of plasma in the back hemisphere of the star. In the case of Betelgeuse, this kind of event happened only once or twice during the last 10 years, with no change in the variability of the star. However, since the great dimming, the variability of Betelgeuse has changed, meaning that this event was such different from the usual mass loss event, that it hustled the dynamics of the photosphere. \

>>>>

Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days, and other shorter periods of 400 and 200 days \citep{kiss_variability_2006}. The origin of the LSP remains poorly known, some tried to explain the LSP by the lifetime of the large convective cells present at the surface of RSG (e.g \cite{stothers_giant_2010}) or by the magnetic field \citep{wood_long_2004}. The shorter periods of 400 and 200 days seem to have a different origin than the LSP. By studying a large sample of RSG, \cite{kiss_variability_2006} attributed the 400 days period to radial pulsation modes, the 200 days period being attributed to the first overtone \citep{joyce_standing_2020}, which was in agreement with the historical work of \cite{stothers_pulsation_1969}.\\

@@

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, it is speculated that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, that might be caused by a mass loss event and leading to this dimming \citep{dupree_great_2022}. Other theories such as a drop in temperature was also mentioned by \cite{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. In this event it turns out that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk.
%due to the presence of large quantities of dust.
Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period, while after the great dimming the main periods of Betelgeuse have shortened, oscillating between 97 days and 230 days \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon. 
Our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 400 and 200 days in  
observational proxies related to the convective activity.\\
%but unrelated to pulsations and which cannot be directly linked to brightness.
%Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
%require,

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes Q.pdf}
    \caption{Same as figure \ref{LS intensity} for stokes $Q$. }
    \label{LS Q}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Stokes U.pdf}
    \caption{Same as figure \ref{LS intensity} for stokes $U$.}
    \label{LS U}
\end{figure}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle linear polarization.pdf}
    \caption{Same as figure \ref{LS intensity} for linear polarization: $\sqrt{Q^2+U^2}$.}
    \label{LS linear polarization}
\end{figure}

@@

Figure \ref{LS Q} and \ref{LS U} show the Lomb-Scargle periodogram of both stokes $Q$ and $U$. It is clear that the periodogram is completely different 
from the intensity one. A strong signal is present in fig. \ref{LS Q} around 200 days. Concerning stokes U in fig. \ref{LS U}, it appears that the main period is around $0.003 \ \mathrm{days^{-1}}$ (330 days). Other period such as the one at 200 days or 250 days ($0.004 \ \mathrm{days^{-1}}$) are present but difficult to trust. From stokes $Q$ and $U$, we recover the 200 days period and the 330 days period is more or less close to the 400 days period found in the literature. 
Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of the total linear polarization of Betelgeuse: $\sqrt{Q^2+U^2}$. 
Interestingly, the periodogram shows a strong power at both 330 days and 200 days, which are the period that we have found in both stokes $Q$ and $U$.

@@

Betelgeuse is observed in average every month by the TBL, and hence
we are able to follow its surface activity through polarimetry imaging. This has  allowed in the past the estimation of  the size and the
 lifetime of the convective cells
at the surface \citep{lopez_ariste_convective_2018}. From these images, we computed a photo-center, which is sensitive to the size and the number of convective cells. 
An homogeneous star will have a photo-center displacement coinciding with barycenter of the star, whereas a star with one or two large convective cells will have a
photo-center displacement more important, up to a few percent of the stellar radius in the case of red supergiants \citep{chiavassa_probing_2022}.

Since the photo-center is linked to surface convection, it is worth checking for periods in its 
dynamics over the 10 years of observations of Betelgeuse at the TBL. Such periods will not 
be completely unrelated to the ones presented in the previous section, but will capture 
other aspects of the phenomena at work.

@@

RSGs encounter mass loss event, where due to stellar winds, plasma is ejected in the interstellar medium. Those events have already been inferred in RSG like $\mu$Cep, where \cite{lopez_ariste_height_2023} found an excess of linear polarization beyond the limb of the star and associated it to rising plumes of plasma in the back hemisphere of the star. In the case of Betelgeuse, this kind of event happened only once or twice during the last 10 years, with no change in the variability of the star. However, since the great dimming, the variability of Betelgeuse has changed, meaning that this event was such different from the usual mass loss event, that it hustled the dynamics of the photosphere. \
